---
title: "PhenoAge Biomarker Range Analysis"
subtitle: "NHANES III data exploration and config generation"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 9, fig.height = 5)
```

# Overview

This report computes normal ranges, plausible ranges, and age-stratified
median defaults for the PhenoAge biological age calculator, using the
NHANES III dataset from the BioAge R package. It writes:

- `config/tests.csv` — updated with data-driven normal and plausible ranges
- `config/defaults.csv` — LOESS-smoothed medians at each integer age (20–84)

## Methodology

1. **Normal ranges** (amber warning): 2.5th–97.5th percentiles from NHANES III
2. **Plausible ranges** (red error): wider of 0.1th–99.9th percentiles and
   clinical case-report extremes, then tightened to catch unit-confusion errors
3. **Age defaults**: raw medians at each integer age, LOESS-smoothed

```{r install-bioage}
if (!requireNamespace("BioAge", quietly = TRUE)) {
  if (!requireNamespace("devtools", quietly = TRUE)) {
    install.packages("devtools")
  }
  devtools::install_github("dayoonkwon/BioAge")
}
library(BioAge)
```

```{r paths}
# When knitting from the analysis/ directory, repo root is one level up
repo_dir <- normalizePath(file.path(getwd(), ".."))
config_dir <- file.path(repo_dir, "config")
cat("Repository root:", repo_dir, "\nConfig directory:", config_dir, "\n")
```

# Load NHANES III

```{r load-data}
data(NHANES3)
cat("NHANES III dataset:", nrow(NHANES3), "observations\n")
cat("Age range:", range(NHANES3$age, na.rm = TRUE), "\n")
```

```{r biomarker-map}
biomarker_map <- data.frame(
  test_id = c("albumin", "creatinine", "glucose", "crp",
              "wbc", "lymphocyte", "mcv", "rcdw", "ap"),
  column = c("albumin_gL", "creat_umol", "glucose_mmol", "crp",
             "wbc", "lymph", "mcv", "rdw", "alp"),
  canonical_unit = c("g/L", "\u00b5mol/L", "mmol/L", "mg/L",
                     "10\u2079 cells/L", "%", "fL", "%", "U/L"),
  to_canonical = c(1, 1, 1, 1, 1, 1, 1, 1, 1),
  stringsAsFactors = FALSE
)
```

# Distribution of each biomarker {.tabset}

Histograms of each biomarker across the full NHANES III cohort, with the
2.5th/97.5th (normal) and 0.1th/99.9th (plausible) percentile boundaries.

```{r compute-ranges}
results <- data.frame(
  test_id = character(),
  normal_low = numeric(), normal_high = numeric(),
  plausible_low = numeric(), plausible_high = numeric(),
  n = integer(),
  stringsAsFactors = FALSE
)

# Store raw values for plotting
biomarker_vals <- list()

for (i in seq_len(nrow(biomarker_map))) {
  tid <- biomarker_map$test_id[i]
  col <- biomarker_map$column[i]
  conv <- biomarker_map$to_canonical[i]

  vals <- NHANES3[[col]]
  if (is.null(vals)) {
    warning(paste("Column", col, "not found in NHANES3"))
    next
  }

  vals <- vals[!is.na(vals)] * conv
  biomarker_vals[[tid]] <- vals
  n <- length(vals)

  normal <- quantile(vals, probs = c(0.025, 0.975))
  extreme <- quantile(vals, probs = c(0.001, 0.999))

  results <- rbind(results, data.frame(
    test_id = tid,
    normal_low = round(as.numeric(normal[1]), 4),
    normal_high = round(as.numeric(normal[2]), 4),
    plausible_low = round(as.numeric(extreme[1]), 4),
    plausible_high = round(as.numeric(extreme[2]), 4),
    n = n,
    stringsAsFactors = FALSE
  ))
}
```

```{r histograms, results='asis', fig.height=4}
for (i in seq_len(nrow(results))) {
  tid <- results$test_id[i]
  unit <- biomarker_map$canonical_unit[biomarker_map$test_id == tid]
  vals <- biomarker_vals[[tid]]

  cat(sprintf("\n## %s\n\n", tid))

  # Trim extreme tails for readable histogram
  qlim <- quantile(vals, probs = c(0.001, 0.999))
  trimmed <- vals[vals >= qlim[1] & vals <= qlim[2]]

  hist(trimmed, breaks = 80, col = "steelblue", border = "white",
       main = paste0(tid, " (", unit, ") — N = ", format(length(vals), big.mark = ",")),
       xlab = unit, ylab = "Frequency")
  abline(v = results$normal_low[i], col = "orange", lwd = 2, lty = 2)
  abline(v = results$normal_high[i], col = "orange", lwd = 2, lty = 2)
  abline(v = results$plausible_low[i], col = "red", lwd = 2, lty = 3)
  abline(v = results$plausible_high[i], col = "red", lwd = 2, lty = 3)
  legend("topright", legend = c("Normal (2.5–97.5%)", "Plausible (0.1–99.9%)"),
         col = c("orange", "red"), lty = c(2, 3), lwd = 2, cex = 0.8)

  cat(sprintf("\n- **N**: %s non-missing\n", format(length(vals), big.mark = ",")))
  cat(sprintf("- **Normal range**: %.3f – %.3f %s\n",
              results$normal_low[i], results$normal_high[i], unit))
  cat(sprintf("- **Plausible range**: %.3f – %.3f %s\n",
              results$plausible_low[i], results$plausible_high[i], unit))
  cat(sprintf("- **Min**: %.4f  **Max**: %.4f  **Median**: %.4f\n\n",
              min(vals), max(vals), median(vals)))
}
```

# Percentile summary

```{r percentile-table}
knitr::kable(results[, c("test_id", "normal_low", "normal_high",
                          "plausible_low", "plausible_high", "n")],
             caption = "NHANES III-derived ranges (canonical units)")
```

# Clinical plausible overrides

```{r clinical-overrides}
clinical_plausible <- list(
  albumin     = c(low = 10,    high = 65),
  creatinine  = c(low = 15,    high = 1800),
  glucose     = c(low = 1.0,   high = 50),
  crp         = c(low = 0.01,  high = 500),
  wbc         = c(low = 0.5,   high = 300),
  lymphocyte  = c(low = 1,     high = 99),
  mcv         = c(low = 50,    high = 140),
  rcdw        = c(low = 9,     high = 35),
  ap          = c(low = 5,     high = 2000)
)

overrides_log <- character()
for (i in seq_len(nrow(results))) {
  tid <- results$test_id[i]
  if (tid %in% names(clinical_plausible)) {
    clin <- clinical_plausible[[tid]]
    old_low <- results$plausible_low[i]
    old_high <- results$plausible_high[i]
    results$plausible_low[i] <- min(results$plausible_low[i], clin["low"])
    results$plausible_high[i] <- max(results$plausible_high[i], clin["high"])
    if (results$plausible_low[i] != old_low || results$plausible_high[i] != old_high) {
      overrides_log <- c(overrides_log, sprintf(
        "%s: widened from [%.4f, %.4f] to [%.4f, %.4f]",
        tid, old_low, old_high, results$plausible_low[i], results$plausible_high[i]))
    }
  }
}

if (length(overrides_log) > 0) {
  cat(paste("- ", overrides_log, collapse = "\n"))
} else {
  cat("No plausible ranges needed widening beyond NHANES extremes.")
}
```

# Unit-confusion tightening

```{r unit-confusion}
convs_csv <- read.csv(file.path(config_dir, "conversions.csv"),
                      stringsAsFactors = FALSE)
conv_list <- split(convs_csv, convs_csv$test_id)

confusion_log <- character()

for (i in seq_len(nrow(results))) {
  tid <- results$test_id[i]
  convs <- conv_list[[tid]]
  if (is.null(convs) || nrow(convs) <= 1) next

  normal_lo <- results$normal_low[i]
  normal_hi <- results$normal_high[i]
  plaus_lo <- results$plausible_low[i]
  plaus_hi <- results$plausible_high[i]

  for (a in seq_len(nrow(convs))) {
    for (b in seq_len(nrow(convs))) {
      if (a == b) next
      fa <- convs$to_canonical_factor[a]
      fb <- convs$to_canonical_factor[b]
      ua <- convs$unit[a]
      ub <- convs$unit[b]

      b_normal_lo_raw <- normal_lo / fb
      b_normal_hi_raw <- normal_hi / fb
      confused_lo <- min(b_normal_lo_raw * fa, b_normal_hi_raw * fa)
      confused_hi <- max(b_normal_lo_raw * fa, b_normal_hi_raw * fa)

      overlap_lo <- max(confused_lo, plaus_lo)
      overlap_hi <- min(confused_hi, plaus_hi)

      if (overlap_lo <= overlap_hi) {
        confusion_log <- c(confusion_log, sprintf(
          "%s: normal %s value (%.2f–%.2f) entered as %s overlaps plausible range",
          tid, ub, b_normal_lo_raw, b_normal_hi_raw, ua))

        if (confused_lo > normal_hi) {
          new_hi <- confused_lo * 0.9
          if (new_hi > normal_hi && new_hi < plaus_hi) {
            confusion_log <- c(confusion_log, sprintf(
              "  -> Tightening plausible_high: %.4f -> %.4f", plaus_hi, new_hi))
            plaus_hi <- new_hi
          }
        }
        if (confused_hi < normal_lo) {
          new_lo <- confused_hi * 1.1
          if (new_lo < normal_lo && new_lo > plaus_lo) {
            confusion_log <- c(confusion_log, sprintf(
              "  -> Tightening plausible_low: %.4f -> %.4f", plaus_lo, new_lo))
            plaus_lo <- new_lo
          }
        }
      }
    }
  }
  results$plausible_low[i] <- round(plaus_lo, 4)
  results$plausible_high[i] <- round(plaus_hi, 4)
}

if (length(confusion_log) > 0) {
  cat(paste("- ", confusion_log, collapse = "\n"))
} else {
  cat("No unit-confusion overlaps detected.")
}
```

# Final ranges

```{r final-ranges}
knitr::kable(results[, c("test_id", "normal_low", "normal_high",
                          "plausible_low", "plausible_high")],
             caption = "Final ranges after clinical overrides and unit-confusion tightening")
```

# Age-stratified defaults {.tabset}

For each biomarker, we compute the raw median at each integer age,
then fit a LOESS smooth. The plots show:

- **Grey dots**: raw medians at each integer age
- **Blue line**: LOESS-smoothed median
- **Orange band**: 2.5th–97.5th percentile at each age (raw)
- **Red dots**: 97.5th percentile at each age

This lets us visually assess whether the LOESS captures real age trends
and whether integer-age resolution is needed vs. a coarser grid.

```{r age-defaults, results='asis', fig.height=5}
age_range <- 20:84
ages_all <- floor(NHANES3$age)

medians_df <- data.frame(age = age_range)

for (i in seq_len(nrow(biomarker_map))) {
  tid <- biomarker_map$test_id[i]
  col <- biomarker_map$column[i]
  conv <- biomarker_map$to_canonical[i]
  unit <- biomarker_map$canonical_unit[i]

  vals <- NHANES3[[col]] * conv

  cat(sprintf("\n## %s\n\n", tid))

  # Raw medians, 2.5th and 97.5th at each integer age
  raw_medians <- sapply(age_range, function(a) {
    v <- vals[ages_all == a & !is.na(vals)]
    if (length(v) >= 5) median(v) else NA
  })
  raw_p025 <- sapply(age_range, function(a) {
    v <- vals[ages_all == a & !is.na(vals)]
    if (length(v) >= 5) quantile(v, 0.025) else NA
  })
  raw_p975 <- sapply(age_range, function(a) {
    v <- vals[ages_all == a & !is.na(vals)]
    if (length(v) >= 5) quantile(v, 0.975) else NA
  })
  raw_n <- sapply(age_range, function(a) {
    sum(ages_all == a & !is.na(vals))
  })

  # LOESS smooth of median
  valid <- !is.na(raw_medians)
  if (sum(valid) >= 10) {
    fit <- loess(raw_medians[valid] ~ age_range[valid], span = 0.4)
    smoothed <- predict(fit, newdata = data.frame(x = age_range))
  } else {
    smoothed <- raw_medians
  }

  medians_df[[tid]] <- round(smoothed, 4)

  # Plot
  ylim <- range(c(raw_p025, raw_p975), na.rm = TRUE)

  plot(age_range, raw_medians, pch = 16, col = "grey50", cex = 0.8,
       xlab = "Age (years)", ylab = paste0(tid, " (", unit, ")"),
       main = paste0(tid, " — median and 95% range by age"),
       ylim = ylim)

  # Shade the 2.5th–97.5th band
  valid_band <- !is.na(raw_p025) & !is.na(raw_p975)
  polygon(c(age_range[valid_band], rev(age_range[valid_band])),
          c(raw_p025[valid_band], rev(raw_p975[valid_band])),
          col = rgb(1, 0.65, 0, 0.15), border = NA)

  # 97.5th percentile points
  points(age_range, raw_p975, pch = 4, col = "red", cex = 0.6)
  points(age_range, raw_p025, pch = 4, col = "red", cex = 0.6)

  # LOESS line
  lines(age_range, smoothed, col = "steelblue", lwd = 2.5)

  legend("topright",
         legend = c("Raw median", "LOESS smooth", "2.5th/97.5th pctile"),
         col = c("grey50", "steelblue", "red"),
         pch = c(16, NA, 4), lty = c(NA, 1, NA), lwd = c(NA, 2.5, NA),
         cex = 0.8, bg = "white")

  # Sample size per age
  cat(sprintf("\n- **Age coverage**: %d of %d integer ages have N >= 5\n",
              sum(raw_n >= 5), length(age_range)))
  cat(sprintf("- **N per age**: min %d, median %d, max %d\n",
              min(raw_n), median(raw_n), max(raw_n)))
  cat(sprintf("- **LOESS range**: %.3f – %.3f %s\n\n",
              min(smoothed, na.rm = TRUE), max(smoothed, na.rm = TRUE), unit))
}
```

# Sample size by age

```{r sample-size, fig.height=4}
n_per_age <- sapply(age_range, function(a) sum(ages_all == a, na.rm = TRUE))
barplot(n_per_age, names.arg = age_range, col = "steelblue", border = "white",
        xlab = "Age (years)", ylab = "N observations",
        main = "NHANES III sample size by integer age")
```

# Write output files

```{r write-defaults}
medians_path <- file.path(config_dir, "defaults.csv")
write.csv(medians_df, medians_path, row.names = FALSE)
cat("Written:", medians_path, "\n")

knitr::kable(head(medians_df, 10), caption = "First 10 rows of defaults.csv")
```

```{r write-tests}
tests_csv <- read.csv(file.path(config_dir, "tests.csv"),
                      stringsAsFactors = FALSE)

tests_csv$normal_low <- NULL
tests_csv$normal_high <- NULL
tests_csv$plausible_low <- NULL
tests_csv$plausible_high <- NULL

tests_out <- merge(tests_csv, results[, c("test_id", "normal_low", "normal_high",
                                           "plausible_low", "plausible_high")],
                   by = "test_id", all.x = TRUE)

tests_out <- tests_out[, c("test_id", "name", "canonical_unit",
                            "normal_low", "normal_high",
                            "plausible_low", "plausible_high")]
tests_out <- tests_out[match(tests_csv$test_id, tests_out$test_id), ]

output_path <- file.path(config_dir, "tests.csv")
write.csv(tests_out, output_path, row.names = FALSE)
cat("Written:", output_path, "\n")

knitr::kable(tests_out, caption = "Final tests.csv", row.names = FALSE)
```
